---
title: "Portfolio 10 -- NBA Playoffs"
author: "Ryan Wheat"
date: "05/08/2023"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, include = FALSE}
library(tidyr)
library(tidyverse)
library(dplyr)
library(haven)
library(lmerTest)
```

```{r load-data, include = FALSE}

RAPTOR <- read_csv("C://Users/12107/OneDrive/Documents/GitHub/Portfolios 2023/Portfolio12/modern_RAPTOR_by_team.csv")

```

## Background

It's cliche in the sports world to joke that the NBA playoffs are a completely different sport from the Regular Season, because the quality of basketball is regarded to be much higher. There are typically two points made to support this. (1) Players generally play much harder when the stakes are higher, and (2) the "great" players tend to rise to the challenge. 

Indeed, the playoffs seem to establish who the best players in the league truly are, because there are many playoff games one could point to in which an NBA legend refused to let their team lose. But in the regular season, the stars tend not to will their teams to victory nearly as often.

These perceived differences in basketball play between the playoffs and the regular season make playoff basketball substantially more fun to watch -- but does empirical evidence support these ideas? That is, do the great players actually perform *better* in the playoffs, consistent with the lay theories? To my knowledge, this question has not faced empirical scrutiny.

To investigate this question, I've downloaded data from fivethirtyeight.com containing their advanced analytics on all NBA players from 2014-present. Specifically, they have a metric known as "RAPTOR," which goes beyond typical box score numbers (e.g., how many points, rebounds, assists, etc. a player gets) to to quantify how "good" a basketball player is. That is, some basketball players play well (or poorly) in ways that do not necessarily show up in traditional box score statistics -- and RAPTOR accounts for these things.

In any case, I am going to split this file up between the regular season and the playoffs, and see if any differences quality of performance emerge as a function of their performance during the regular season.

## Data Cleaning

First, I need to filter the data so that each player has both (1) data from the regular season, and (2) data from the playoffs for a given NBA season.

```{r wrangling}

#sort data so that each player included has data from the regular season and playoffs for a given year

RAPTOR_filtered <- RAPTOR %>%
  group_by(player_id, season) %>%
  mutate(filter_by = if_else(length(season) > 1 & length(season) <= 2, 1, 0)) %>%
  ungroup()

RAPTOR_filtered <- RAPTOR_filtered %>%
  group_by(player_id, season) %>%
  filter(filter_by == 1) %>%
  ungroup()

#there are some people who were on multiple teams in a given season that did not make the playoffs. Let's filter those players out.

RAPTOR_filtered <- RAPTOR_filtered %>%
  group_by(player_id, season) %>%
  filter("PO" %in% season_type) %>%
  ungroup()

RAPTOR_filtered %>%
  count(season_type)

```

Okay, we've now got a dataset of players who played for a given team in a given year during both the regular season and the playoffs.

Let's now do a median split on total_RAPTOR to seperate the "good" players from the "bad" players. We'll also seperate by +-1 standard deviation on total_RAPTOR. 

```{r splitting}

#regular season and playoff data

RAPTOR_rs <- RAPTOR_filtered %>%
  filter(season_type == "RS")

RAPTOR_po <- RAPTOR_filtered %>%
  filter(season_type == "PO")

#median split

RAPTOR_median_rs <- RAPTOR_rs %>%
  mutate(median_split = case_when(raptor_total > median(raptor_total) ~ "Good Player",
                                  raptor_total < median(raptor_total) ~ "Bad Player",
                                  raptor_total == median(raptor_total) ~ "Average Player"))

RAPTOR_median_po <- RAPTOR_po %>%
  mutate(median_split = case_when(raptor_total > median(raptor_total) ~ "Good Player",
                                  raptor_total < median(raptor_total) ~ "Bad Player",
                                  raptor_total == median(raptor_total) ~ "Average Player"))


#+/- 1 SD split

raptor_mean <- mean(RAPTOR_filtered$raptor_total)
raptor_sd <- sd(RAPTOR_filtered$raptor_total)

RAPTOR_SD_rs <- RAPTOR_rs %>%
  mutate(sd_split = case_when(raptor_total > raptor_mean + raptor_sd ~ "Good Player",
                              raptor_total < raptor_mean - raptor_sd ~ "Bad Player",
                              .default = "Average Player"))

RAPTOR_SD_po <- RAPTOR_po %>%
  mutate(sd_split = case_when(raptor_total > raptor_mean + raptor_sd ~ "Good Player",
                              raptor_total < raptor_mean - raptor_sd ~ "Bad Player",
                              .default = "Average Player"))

RAPTOR_median_rs %>%
  count(median_split)

#put the regular season evaluations into the playoff datasets

RAPTOR_median_rs %>%
  arrange(player_id)

RAPTOR_SD_rs %>%
  arrange(player_id)

RAPTOR_po <- RAPTOR_po %>%
  arrange(player_id) %>%
  mutate(reg_median_split = RAPTOR_median_rs$median_split) %>%
  mutate(reg_sd_split = RAPTOR_SD_rs$sd_split) %>%
  mutate(reg_raptor = RAPTOR_SD_rs$raptor_total)

```

## Main Analyses

Ok, now that we've set that up, I'm going to run a series of regression models. If I were doing a full thing for this, I would probably run something to account for potential lack of independent observations (e.g., players nested within teams). But I'm not going to do all that today.

Instead, I will see if the dichotomous variable of season_type predicts total_raptor. In other words: do players generally play better or worse in the playoffs?

```{r playoffs}

#create variable to show that the same person is 

po_v_rs <- lm(raptor_total ~ season_type, RAPTOR_filtered)

summary(po_v_rs)

```

The lack of signficant results suggests that the playoffs vs. regular season distinction doesn't matter. That is, players do not generally play better or worse in the playoffs than they do in the regular season.

Next, I will run a regression model to see whether each player's playoff total_raptor can be predicted by their regular season total_raptor. In other words, does how well a player plays in the regular season predict how well they play in the playoffs? If RAPTOR is a valid measure of player quality, I expect this to be a strong relationship.

```{r regular-season-predictor}

reg_po_raptor <- lm(raptor_total ~ reg_raptor, RAPTOR_po)

summary(reg_po_raptor)

```

Good -- the higher a player's regular season RAPTOR, the higher their postseason RAPTOR in a given year. You would hope this would be the case if a metric such as RAPTOR was supposed to be a good measure of how good a basketball player was.

Now, I'm going to create a difference score between postseason and regular season RAPTOR to see if being good (and bad) in the regular season predicts *change* in how good (or bad) a player is in the playoffs.

Positive values on this metric indicate that a player got BETTER from the regular season to the playoffs. Negative values indicate that a player got WORSE, instead.

There are two competing hypotheses, here. One possibility is that being bad in the regular season will predict improving in the playoffs, because of a floor effect. This seems unlikely to me. The other possibility is that being good (or bad) in the regular season predicts getting better (or worse) in the playoffs, consistent with lay conceptions.

```{r change-model}

#create variable

RAPTOR_po <- RAPTOR_po %>%
  mutate(raptor_change = raptor_total - reg_raptor)

po_change_m <- lm(raptor_change ~ reg_median_split, RAPTOR_po)

po_change_sd <- lm(raptor_change ~ reg_sd_split, RAPTOR_po)

summary(po_change_m)
summary(po_change_sd)

```

This is interesting. "Good players," identified as those above the median in RAPTOR, seem to get worse in the playoffs. "Bad players," or those below the median, seem to improve.

This also seems to be the case if doing a split at +/- 1 SD (instead of the median). That is, "good" players are predicted to have worse RAPTOR in the playoffs than the regular season compared to average players, and "bad" players are predicted to have a better raptor in the playoffs than they did in the regular season compared to average players.

There are a lot of things that could be happening here, and I'm sure there is a way to unpack the data that better answers the question of interest. There are also a number of limitations to the analyses I've run, especially regarding the way I've operationalized a "good" basketball player for this portfolio.

However, taking these results at face value, the empirical evidence I've gathered goes against popular wisdom: good players don't seem to systematically improve in the NBA playoffs. In fact, they seem to get worse.
